---
title: "Homework 1"
author: "Sarosh Farhan (24210969)"
format: 
  pdf:
    include-in-header: 
      text: |
        \usepackage{xcolor}
editor: visual
execute:
  echo: false
---

# Introduction

The \textcolor{blue}{Iris dataset} is a classic dataset in pattern recognition and multivariate statistics. It contains measurements of *sepal length*, *sepal width*, *petal length*, and *petal width* for 150 flowers from three species of iris: *setosa*, *versicolor*, and *virginica*.

# Approach

I will explore the distribution of petal length versus petal width for each species using a scatter plot. This will help me visualize how distinct the species are based on these two features.

# Statistics

The following table provides the count, mean, and standard deviation of \textcolor{purple}{petal length} and \textcolor{orange}{petal width} for each iris species.

```{r, error=FALSE, warning=FALSE, message=FALSE}
library(dplyr)
library(knitr)

# Summary statistics
summary_table <- iris %>%
  group_by(Species) %>%
  summarise(
    Count = n(),
    MeanPetalLength = mean(Petal.Length),
    SDPetalLength = sd(Petal.Length),
    MeanPetalWidth = mean(Petal.Width),
    SDPetalWidth = sd(Petal.Width)
  )

# Display as a nice table
kable(summary_table, digits = 2, caption = "Summary Statistics by Species")
```

## Summary table inferences

\textcolor{red}{Setosa} has the ***smallest petals*** overall, with an average petal length of **1.46 cm** and width of **0.25 cm**. The low standard deviations indicate that \textcolor{red}{Setosa} flowers are consistently small in size.

\textcolor{green}{Versicolor} shows ***moderate petal*** size, averaging **4.26 cm** in length and **1.33 cm** in width. It also exhibits slightly higher variation compared to \textcolor{red}{Setosa}.

\textcolor{blue}{Virginica} has the ***largest petals***, with mean petal dimensions of **5.55 cm** (length) and **2.03 cm** (width). It also displays the greatest variability, especially in petal width.

These differences are **substantial and consistent**, making **petal length and width excellent predictors** for classifying iris species.

# Plot

The scatter plot below shows the relationship between petal length and petal width, colored by species:

```{r, fig.align='center', fig.width=8, fig.height=5}
#| label: fig-scatter
#| fig-cap: "Scatter plot of Petal Length vs Petal Width by Species"
#| fig-align: center

library(ggplot2)
# scatter plot as per petal width and petal length (cm)
ggplot(iris, aes(x = Petal.Length, y = Petal.Width, color = Species)) +
  geom_point(size = 2) +
  theme_minimal() +
  labs(
    title = "Petal Length vs Petal Width by Species",
    x = "Petal Length (cm)",
    y = "Petal Width (cm)"
  ) +
  theme(
    # Centering the title
    plot.title = element_text(hjust = 0.5)  
  )
```

## Interpretation of the plot

From @fig-scatter we can infer the following:

\textcolor{red}{Setosa} is clearly separable, its petals are significantly smaller in both length and width compared to other two species.

\textcolor{green}{Versicolor} lies in the middle range for both dimensions.

\textcolor{blue}{Virginica} has the largest petals, both in length and width.

There is a visible positive correlation between petal length and petal width within each species.The species form distinct clusters, particularly \textcolor{red}{Setosa}, which is completely separated from the others. This makes petal measurements powerful features for classification tasks like species prediction.
